Clone and graft:\\ Testing scientific applications as they are built

### 2015-01-21 turcksin-spring-mechanics/paper.tex
Go through it one more time.

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
\marginpar{After documenting spring.py, split it and simplify the first
  version to the pieces we need here}
\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]{spring.py}

\begin{wrapfigure}{R}{0.2\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.18\textwidth]{tests/testcase-1.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Trajectories for the two balls with the input file from
    Section~\ref{sec:first-steps}. Note that axes are not shown at the same
    scale.}
  \vspace*{-3mm}
  \label{fig:testcase-1}
\end{wrapfigure}

@@

We often have an intuitive idea of how to do the former. For example, we could
note that the center of mass at the end time $T=5$ should have fallen a
distance of $\delta z=\frac 12 g T^2=\frac{9.81 \cdot 25}{2}\textrm m\approx
123\textrm{m}$, well in agreement with what Fig.~\ref{fig:testcase-1} shows. We
could also compare with another code. Or we could look at special cases for
which we have 
an analytic solution. For our model problem, one could for example prescribe
initial velocities so that the bodies are at rest and initial positions so
that the two bodies are separated by the rest length of the spring; the two
bodies then simply fall straight down and we can easily compute on a piece of
paper where they should be at time $t$. The following input file
(\texttt{testcase-2.json}) does this:
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-2.json}

In the end, verifying whether the program is correct is problem dependent. Let
us not go too deep into this because the second answer -- how to ensure that
functionality continues to work -- is actually much more important in practice
even though few people realize this when they start developing software. The
issue essentially boils down to this: I need mechanisms that ensure that
functionality if I have verified today works correctly, does not break when
implementing new functionality. This is necessary because good programmers
understand that they will make mistakes; what makes them good programmers is
that they develop ways to minimize the impact by helping them find problems
quickly.

There is a lot of research on bugs in programs. Fundamentally, the essence of
this research is that it cannot be avoided altogether when programming. On the
other hand, the best programs are those that make it easy to spot bugs and fix
them quickly. The worst programs are those that are not tested; an empirically
true statement is that programs that are not tested produce wrong answers --
where the emphasis is on the fact that they \textit{produce} wrong answers,
not that they \textit{may produce} wrong answers. Experienced programmers all
attest to this statement. Only bad or inexperienced programmers will assume
that they do not introduce new bugs in old parts of their programs.%
\footnote{In fact, what seems to set good programmers apart more than many
  other factor is not that they create fewer bugs -- which they do, though not
  dramatically so -- but that they understand that they create bugs and that
  they change their behavior to accommodate for it, for example by developing
  a habit of rigorous testing that helps them find these bugs quickly.}

The importance of testing relies in the cost of fixing problems. If we keep
implementing new functionality, eventually some old feature will stop
working. If we have no tests that detect this quickly, this may go unnoticed
for a while but eventually we will try to re-run an old computation with the
new version of the code and it will not work any more. So where to start? What
have we changed in the intervening months that may cause this? There is likely
a lot of new code we have written since then, and we may have forgotten about
a fair share of the details since then. It will take a long time to figure out
where exactly the problem is. On the other hand, if we had a testsuite that we
run after essentially every change, we will be notified that something stopped
working immediately -- with a relatively small number of changes made since
the test last worked, and everything still fresh on our mind, finding the bug
should be a much quicker issue.

So how do we achieve this? The general idea is that we save the test cases we
have previously used to convince ourselves that a new part of the program was
correct. To make this work, we have to have three things:
\begin{enumerate}
\item The input file for a previously verified testcase.
\item The output when running the code on this input file; having verified the
  feature, we know or at least believe that this output is correct.
\item A way to run a new version of the code on the output file, to compare
  its current output against the stored output, and to raise a flag if they
  are not the same.
\end{enumerate}
The first two are really not very complicated -- create a directory
\texttt{tests/} in your project and create two files in it, say
\texttt{testcase-1.json} and \texttt{testcase-1.output}. The last part can
be as easy as running commands such as
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
  python masses-and-springs.py testcase-1.json > testcase-1.output.new
  diff testcase-1.output testcase-1.output.new
\end{lstlisting}
You would repeat all of this for \texttt{testcase-2.json}, of course.
If the \texttt{diff} program produces no output, then the two files are the
same and your program computed the same answer on this input file as it did
back when you created the \texttt{testcase-1.output} file. If there is a
difference, then you know that you just broke one of the testcases. If you
purposefully changed it -- for example because you changed the output format,
or because you fixed behavior you know was erroneous and have verified that it
is now correct, then copy the new output over the new one. If you did not
purposefully change it, then you know it's time to break out the debugger.

@@

We write tests because we expect that our code base will continue to change --
to implement new functionality, or to fix existing bugs. So let's do this:
assume we now also want to include air friction on the two balls, and we will
use that the air friction force is proportional to the square of the speed and
in the opposite direction of the velocity \cite{...}. Then our model
\eqref{eq:ode} needs to change to the following:
\begin{align}
  \label{eq:ode2}
  m_i 
  \mathbf x_i''(t)
  &=
  m_i \mathbf g
  -
  D \left(\|\mathbf x_2(t) - \mathbf x_1(t)\| - L\right) \mathbf d_i
  \underbrace{-C_i \; \|\mathbf x_i(t)\| \; \mathbf x_i(t)}_{\text{friction force}}.
\end{align}
The new term contains constants $C_i$ that denote air friction coefficients of
the two balls. These coefficients can be computed from the cross section of
the ball, their surface roughness, the viscosity of the air, and other
factors, but this is not important here -- we will simply read them from the
input file.

If we now require that every input file contains an array of friction
coefficients $C_i$, we can no longer use old input files. A better strategy is
to define default values for all non-required parameters and overwrite them
with the values in the input file if specified. A backward compatible choice
of input parameter for the $C_i$ is to give them a default value of zero
because in that case equation \eqref{eq:ode2} equals the previous version,
\eqref{eq:ode}.

Implementing all of this leads to the following program:
\lstinputlisting[frame=single,basicstyle=\footnotesize,language=Python]{spring.py}

@@

Next, we need to write new tests for the new functionality. This is where the
\textit{clone and graft} technique finally comes into play: take one of the
existing tests (i.e., \textit{clone} it) and add to it the minimal number of
parameters to exercise the new functionality (i.e., \textit{graft} onto
it). So let's take the simpler of the two previous testcases,
\texttt{tests/testcase-2.json}, copy it to \texttt{tests/testcase-3.json} and
add to it something about air friction: 
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-3.json}
To verify that the results we get are correct, we can use that in this test
case the masses and friction coefficients of the two balls are the same, the
two balls are separated by a distance equal to the rest length of the spring,
and they start at rest. Then they will simply fall as if there was no spring,
and we know that they will approach a terminal speed $v=\sqrt{mg/C}$ where
downward gravity and upward air friction cancel. For the current case, this
would be $v=\sqrt{9.81}$ because we have chosen $m=1, g=9.81, C=1$.

Is one new testcase enough? The answer is rarely yes. For example, it is easy
to make indexing errors and choosing masses and friction coefficients equal
would not allow us to detect errors where, for example, we used
\texttt{friction[1]} in the code where we intended to use \texttt{friction[i]}.
So let's copy \texttt{tests/testcase-3.json} to
\texttt{tests/testcase-4.json}:
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
Here, the second ball has a greater mass, but the same air friction
coefficient (think of it as denser but with the same diameter). How do we know
whether our code is correct for this case as well? This may be more difficult
to establish, but it isn't always necessary. For example, we could look at the
output and realize that the second ball falls faster and, because it is
connected to the first by a spring, then starts to drag the first behind it as
it falls. Such qualitative verification is sometimes the best one can do. At
the same time, while it may not guarantee that the code is indeed correct, it
will nonetheless serve as a useful test in the future to ensure that existing
functionality does not break as a result of further development -- in the long
term the more important of the two reasons to write tests.

Because it is so easy to create tests, it may be worthwhile cloning
\texttt{tests/testcase-3.json} a second time into
\texttt{tests/testcase-5.json} where now we just set the friction coefficients
differently:
\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
Now the two balls have the same mass but the second one has a higher drag
coefficient (think of the first ball as smooth and the second one as covered
in fur). Our ``smell check'' would now be to verify that the second ball drags
behind the first.

@@

But this does not always work. There are cases where you will have to write a
new test from scratch. To give an example of where this was necessary,
consider \aspect{}: imagine that at the beginning, it could only compute
convection in a box. We would have testcases that described different
temperature boundary conditions on this box, and various other things. But
when implementing a spherical geometry, none of these testcases quite fit:
they all had to say what the temperature is on the six faces of the box, they
made the implicit assumption that gravity points downward instead of inward,
etc. It was time to start with a new testcase. If, at a later time you then
clone and graft from this new testcase, then it becomes the root of a separate
tree (let's call it an \textit{exemplar} testcase) and the inheritance graph
will turn into a \textit{forest graph} (a collection of tree graphs).

\begin{wrapfigure}{L}{0.45\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[height=0.45\textwidth,angle=-90]{figures/graph.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Connection graph of the 144 tests that are part of the
    \aspect{} code base.}
  \vspace*{-3mm}
  \label{fig:aspect-tests}
\end{wrapfigure}
Finally, there are cases where you copy-paste pieces from multiple existing
tests into a new one, for example because you want to test the combination of
features for which you have already written individual testcases. At this
point, the metaphor of inheritance graphs starts to break down. However, in
practice, most tests are still closely related to some others. A visualization
of this for the 144 tests that are part of \aspect{}, each represented by a
dot, is shown in Fig.~\ref{fig:aspect-tests} where we draw tests farther apart
of there are more textual differences between the input files. Many tests come
in groups of 2, 3 or 4 that only differ in a few lines of input -- simple
modifications such as those between testcases 3, 4 and 5 above. On the other
hand, other group are further apart and are derived from different
examplars.

@@

In practice, once you have accumulated a number of testcases, you will
encounter practical difficulties. First, when you try to find the right
starting point to clone and graft a new testcase, you will not recall if
\texttt{tests/testcase-3.json} or \texttt{tests/testcase-5.json} is the one
that's closest to the situation you have in mind (for example, if you want to
augment your model and you are looking again for a testcase with equal masses
and friction coefficients).

@@

In practice, you're not likely to quickly get to a project size where learning
to use tools like \texttt{CTest} and \texttt{CDash} are important
considerations. In practice, the point where using such tools becomes
indispensible is when your code has several 10,000 lines of code or more --
several years worth of work. When you get to this point, your project will
have all sorts of other growing pains as well, such as having to deal with
external users, running mailing lists, portability, managing complexity and
other issues. Some of these are discussed in \cite{BH13}. The point simply is
that there is no need to worry too much about testing tools for programs of
this size: care about writing tests from the beginning and if the number of
tests becomes too unwieldy, there will be professional-grade tools you can
switch to that can manage this for you. In the meantime, a little shell script
such as the following will do just fine:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
  # loop over all .json input files
  for jsonfile in *.json ; do
    # execute our program on it and pipe the output into a file
    python ${jsonfile} > ${jsonfile}.new

    # then compare output; only ask 'diff' to report whether the
    # files are the same or not, and create output depending on this:
    # print an 'X' if there is a difference, a '.' if there isn't. this
    # ensures that test failures are easily visible
    if (diff ${jasonfile} ${jsonfile}.new > /dev/null) ; then
      echo " .    ${jasonfile}" ;
    else 
      echo " X    ${jasonfile}" ;
    fi
  done
\end{lstlisting}
If you call this script, say, \texttt{tests/compare.sh}, then running it using
the command \texttt{cd tests/ ; bash -e compare.sh} should produce output like
this:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
 .    testcase-1.json
 .    testcase-2.json
 .    testcase-3.json
 .    testcase-4.json
 .    testcase-5.json
\end{lstlisting}
On the other hand, if your output looks like this
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
 .    testcase-1.json
 .    testcase-2.json
 .    testcase-3.json
 X    testcase-4.json
 .    testcase-5.json
\end{lstlisting}
then you know that you just broke some functionality that
\texttt{testcase-4.json} checks. Given how easy it is to run this script, you
want to run this script often to find new bugs as early as possible.

@@

From experience, it turns out that this is not often a real problem. The vast
majority of patches concern just small, isolated pieces of the software and in
those cases, only one or a few testcases will start failing. It is then easy
to identify the simplest one of these (assuming that they were given useful
names, see above in Section~\ref{sec:many-tests}), and that is the one one
will then use for debugging. If one breaks fundamental functionality somewhere
in the core of the software, a large number of tests will start to fail
but in those cases you will either be able to quickly find out what is
happening, or be able to debug the problem by choosing \textit{any} simple
test that was among those that broke. In summary, from a practical
perspective, we have found that it is not usually a problem to have testcases
that check overlapping parts of the software.

@@

At the base of the problem is the fact that floating point numbers do not
enjoy the same mathematical properties as integers (or, in fact, as real
numbers). For example, if you take into account round-off, then $(a+b)+c$ is in
general \textit{not} the same as $a+(b+c)$ or $(a+c)+b$. You can try this out
with $a=1.0, b=-1.0, c=1e-20$: in double precision, you will get
$(a+b)+c=1e-20, a+(b+c)=(a+c)+b=0$. This matters because if you make changes
to the code base, upgrade the compiler, switch from one processor or machine
to another, or change compiler optimization level, floating point operations
will be executed in different order and the results will be different at the
level of roundoff. In other words, your tests will start to fail even though
nothing fundamentally changed -- but the output will be different.

@@

A second area that creates problems that codes often try to compute zeros, but
because of roundoff what is actually produced are numbers of the kind
4.863e-19. A human recognizes that this is a zero and, consequently, equal to
2.489e-19 if all of the numbers around it are on the order of one. But the
\texttt{diff} command will complain, and consequently signal a difference.

The solution to these problems that we have ultimately come to use
is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
\texttt{diff} but it recognizes if the text it compares in two files is
actually a number. If two numbers differ by less than some relative or
absolute tolerance, then they are considered equal even if the characters by
which they are represented in an output file are different. For example, you
would replace the \texttt{diff} command used previously by the
following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
10^{-6}$ or $|a-b|\le 10^{-8} \min\{|a|,|b|\}$:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
\end{lstlisting}
Using this trick, numbers that are written into the output file
\texttt{testcase-1.output.new} on one machine with slight differences to the
ones originally saved in \texttt{testcase-1.output} will still be considered
the same, and the comparison will succeed.

@@

When we talk to students who are just starting writing their own codes, then
they typically understand the importance of testing in some abstract
sense. But almost all of them have two fundamental misconceptions: First, that
their codes are either written well enough or will not be changed in ways
fundamental enough to suggest that these changes might break existing
functionality. And second, that doing adequate testing is just too hard and
cumbersome to be a justifiable expense of time.

>>>>

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
import sys
import json
import numpy

# Read the name of the input file from the command line, and read options from the file:
assert len(sys.argv) == 2, 'Please provide an input file.'
with open(sys.argv[1], 'r') as f:
    settings = json.loads(f.read())

# Then retrieve the various parameters from what we just read:
D = settings['spring constant']
L = settings['spring rest length']
x1_0 = settings['initial position'][0]
x2_0 = settings['initial position'][1]
v1_0 = settings['initial velocity'][0]
v2_0 = settings['initial velocity'][1]

m = [settings['masses'][0], settings['masses'][1]]

# describe the differential equation as a first order ODE:
y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
      v1_0[0], v1_0[1], v1_0[2], v2_0[0], v2_0[1], v2_0[2]]

def f(t, y):
    p1 = y[0:3]
    p2 = y[3:6]
    v1 = y[6:9]
    v2 = y[9:12]

    g = [0., 0., -9.81]

    dist = numpy.linalg.norm(p2-p1)
    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
    return numpy.concatenate([v1, v2, a1, a2])

# Next create an object that can integrate the ODE numerically:
start_time = 0.
end_time   = 5
import scipy.integrate
integrator = scipy.integrate.ode(f)
integrator.set_integrator('vode', rtol=1e-6)
integrator.set_initial_value(y0, start_time)

# With this, do the integration step by step, appending values to an array in each step:
t_values = [start_time]
y_values = numpy.array([y0])
while integrator.successful() and integrator.t < end_time:
    integrator.integrate(end_time, step=True)
    t_values.append(integrator.t)
    y_values = numpy.vstack((y_values, integrator.y))

# Having done so, output the number of time steps and the final positions:
print "time steps:", len(t_values)
print "final position:", y_values[len(t_values)-1,0:3], y_values[len(t_values)-1,3:6]
\end{lstlisting}

\begin{wrapfigure}{R}{0.3\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.28\textwidth]{figures/testcase-1.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Trajectories for the two balls with the input file from
    Section~\ref{sec:first-steps}. Note that axes are not shown at the same
    scale.}
  \vspace*{-3mm}
  \label{fig:testcase-1}
\end{wrapfigure}

@@

We often have an intuitive idea of how to do the former. For example, we could
note that the center of mass at the end time $T=5$ should have fallen a
distance of $\delta z=\frac 12 g T^2=\frac{9.81 \cdot 25}{2}\textrm m\approx
123\textrm{m}$, well in agreement with what Fig.~\ref{fig:testcase-1} shows. We
could also compare with another code. Or we could look at special cases for
which we have 
an analytic solution. For our model problem, one could for example prescribe
initial velocities so that the bodies are at rest and initial positions so
that the two bodies are separated by the rest length of the spring; the two
bodies then simply fall straight down and we can easily compute on a piece of
paper where they should be at time $t$. The following input file
(\texttt{tests/testcase-2.json}) does this:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-2.json}

In the end, verifying whether the program is correct is problem dependent. Let
us not go too deep into this because the second answer -- how to ensure that
functionality continues to work -- is actually much more important in practice
even though few people realize this when they start developing software. The
issue essentially boils down to this: We need mechanisms that ensure that
the functionality we have verified works correctly today, does not break when
we implement new functionality tomorrow. This is necessary because good programmers
understand that they will make mistakes; what makes them good programmers is
that they develop ways to minimize the impact by helping them find problems
quickly.

There is a lot of research on bugs in programs. Fundamentally, the essence of
this research is that it cannot be avoided altogether when programming. On the
other hand, the best programs are those that make it easy to spot bugs and fix
them quickly. The worst programs are those that are not tested; an empirically
true statement is that programs that are not tested produce wrong answers --
where the emphasis is on the fact that they \textit{do} produce wrong answers,
not that they \textit{may produce} wrong answers. Experienced programmers all
attest to this statement. Only bad or inexperienced programmers will assume
that they do not introduce new bugs in old parts of their programs.%
\footnote{In fact, what seems to set good programmers apart more than many
  other factor is not that they create fewer bugs -- which they do, though not
  dramatically so -- but that they understand that they create bugs and that
  they change their behavior to accommodate for it, for example by developing
  a habit of rigorous testing that helps them find these bugs quickly.}

The importance of testing relies in the cost of fixing problems. If we keep
implementing new functionality, eventually some old feature will stop
working. If we have no tests that detect this quickly, this may go unnoticed
for a while but eventually we will try to re-run an old computation with the
new version of the code and it will not work any more. So where to start? What
have we changed in the intervening months that may cause this? There is likely
a lot of new code we have written since then, and we may have forgotten about
a fair share of the details since then. It will take a long time to figure out
where exactly the problem is. On the other hand, if we had a testsuite that we
run after essentially every change, we will be notified immediately that something stopped
working -- with a relatively small number of changes made since
the test last worked, and everything still fresh on our mind, finding the bug
should be a much quicker issue.

So how do we achieve this? The general idea is that we save the test cases we
have previously used to convince ourselves that a new part of the program was
correct. To make this work, we have to have three things:
\begin{enumerate}
\item The input file for a previously verified testcase.
\item The output when running the code on this input file; having verified the
  feature, we know or at least believe that this output is correct.
\item A way to run a new version of the code on the output file, to compare
  its current output against the stored output, and to raise a flag if they
  are not the same.
\end{enumerate}
The first two are really not very complicated -- create a directory
\texttt{tests/} in your project and create two files in it, say
\texttt{testcase-1.json} and the corresponding output
\texttt{testcase-1.output}. When it comes time to test at a later time, step 3
above can be as easy as running commands such as
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
  python masses-and-springs.py testcase-1.json > testcase-1.output.new
  diff testcase-1.output testcase-1.output.new
\end{lstlisting}
You would repeat all of this for \texttt{testcase-2.json}, of course.
If the \texttt{diff} program produces no output, then the two files are the
same and your program computed the same answer on this input file as it did
back when you created the \texttt{testcase-1.output} file. If there is a
difference, then you know that you just broke one of the testcases. If you
purposefully changed it -- for example because you changed the output format,
or because you fixed behavior you know was erroneous and have verified that it
is now correct, then copy the new output over the new one. If you did not
purposefully change what the program computes, then you know it's time to break out the debugger.

@@

We write tests because we expect that our code base will continue to change --
to implement new functionality, or to fix existing bugs. So let's do this:
assume we now also want to include air friction on the two balls, and we will
use that the air friction force is proportional to the square of the speed and
in the opposite direction of the velocity. Then our model
\eqref{eq:ode} needs to change to the following:
\begin{align}
  \label{eq:ode2}
  m_i 
  \mathbf x_i''(t)
  &=
  m_i \mathbf g
  -
  D \left(\|\mathbf x_2(t) - \mathbf x_1(t)\| - L\right) \mathbf d_i
  \underbrace{-C_i \; \|\mathbf x_i(t)\| \; \mathbf x_i(t)}_{\text{friction force}}.
\end{align}
The new term contains constants $C_i$ that denote air friction coefficients of
the two balls. These coefficients can be computed from the cross section of
the ball, their surface roughness, the viscosity of the air, and other
factors, but this is not important here -- we will simply read them from the
input file.

If we now require that every input file contains an array of friction
coefficients $C_i$, we can no longer use old input files. A better strategy is
to define default values for all non-required parameters and overwrite them
with the values in the input file if specified. A backward compatible choice
of input parameters for the $C_i$ is to give them a default value of zero
because in that case equation \eqref{eq:ode2} equals the previous version,
\eqref{eq:ode}.

Implementing all of this leads to the following extension of the program shown
above%
\footnote{This is the file \texttt{spring.py} in the directory corresponding to
this paper}, with the new parts in lines 21--29 and 45--46:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}

@@

Next, we need to write new tests for the new functionality. This is where the
\textit{clone and graft} technique finally comes into play: take one of the
existing tests (i.e., \textit{clone} it) and add to it the minimal number of
parameters to exercise the new functionality (i.e., \textit{graft} onto
it). So let's take the simpler of the two previous testcases,
\texttt{tests/testcase-2.json}, copy it to \texttt{tests/testcase-3.json} and
add to it something about air friction: 
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-3.json}
To verify that the results we get are correct, we can use that in this test
case the masses and friction coefficients of the two balls are the same, the
two balls are separated by a distance equal to the rest length of the spring,
and they start at rest. Then they will simply fall as if there was no spring,
and we know that they will approach a terminal speed $v=\sqrt{mg/C}$ where
downward gravity and upward air friction cancel. For the current case, this
would be $v=\sqrt{9.81}$ because we have chosen $m=1, g=9.81, C=1$.

Is one new testcase enough? The answer is rarely yes. For example, it is easy
to make indexing errors and choosing masses and friction coefficients equal
would not allow us to detect errors where, for example, we used
\texttt{friction[1]} in the code where we intended to use \texttt{friction[i]}.
So let's copy \texttt{testcase-3.json} to
\texttt{testcase-4.json}:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-4.json}
Here, the second ball has a greater mass, but the same air friction
coefficient (think of it as denser but with the same diameter). How do we know
whether our code is correct for this case as well? This may be more difficult
to establish, but it isn't always necessary. For example, we could look at the
output and realize that the second ball falls faster and, because it is
connected to the first by a spring, then starts to drag the first behind it as
it falls. Such qualitative verification is sometimes the best one can do. At
the same time, while it may not guarantee that the code is indeed correct, it
will nonetheless serve as a useful test in the future to ensure that existing
functionality does not break as a result of further development -- in the long
term the more important of the two reasons to write tests.

Because it is so easy to create tests, it may be worthwhile cloning
\texttt{testcase-3.json} a second time into
\texttt{testcase-5.json} where now we just set the friction coefficients
differently:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-5.json}
Now the two balls have the same mass but the second one has a higher drag
coefficient (think of the first ball as smooth and the second one as covered
in fur). Our ``smell check'' would now be to verify that the second ball drags
behind the first.

If we're reasonably convinced that the output is correct, then save the output
in files \texttt{tests/testcase-3.output} through
\texttt{tests/testcase-5.output} for use in future testing.

@@

But this does not always work. There are cases where you will have to write a
new test from scratch. To give an example of where this was necessary,
consider \aspect{}: Imagine that at the beginning, it could only compute
convection in a box. We would have testcases that described different
temperature boundary conditions on this box, and various other things. But
when implementing a spherical geometry, none of these testcases quite fit:
they all had to say what the temperature is on the six faces of the box, they
made the implicit assumption that gravity points downward instead of inward,
etc. It was time to start with a new testcase. If, at a later time you then
clone and graft from this new testcase, then it becomes the root of a separate
tree (let's call it an \textit{exemplar} testcase) and the inheritance graph
will turn into a \textit{forest graph} (a collection of tree graphs).

\begin{wrapfigure}{L}{0.45\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[height=0.45\textwidth,angle=-90]{figures/graph.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Connection graph of the 144 tests that are part of the
    \aspect{} code base.}
  \vspace*{-3mm}
  \label{fig:aspect-tests}
\end{wrapfigure}
Finally, there are cases where you copy-paste pieces from multiple existing
tests into a new one, for example because you want to test the combination of
features for which you have already written individual testcases. At this
point, the metaphor of inheritance graphs starts to break down. However, in
practice, most tests are still closely related to some others. A visualization
of this for the 144 tests that are part of \aspect{}, each represented by a
dot, is shown in Fig.~\ref{fig:aspect-tests} where we draw tests farther apart
if there are more textual differences between the input files. Many tests come
in groups of 2, 3 or 4 that only differ in a few lines of input -- simple
modifications such as those between testcases 3, 4 and 5 above. On the other
hand, other group are further apart and are derived from different
examplars.

@@

In practice, once you have accumulated a number of testcases, you will
encounter practical difficulties. First, when you try to find the right
starting point to clone and graft a new testcase, you will not recall if
\texttt{testcase-3.json} or \texttt{testcase-5.json} is the one
that's closest to the situation you have in mind (for example, if you want to
augment your model and you are looking again for a testcase with equal masses
and friction coefficients).

@@

In practice, you're not likely to quickly get to a project size where learning
to use tools like \texttt{CTest} and \texttt{CDash} are important
considerations. In practice, the point where using such tools becomes
indispensible is when your code has several 10,000 lines of code or more --
several years worth of work. When you get to this point, your project will
have all sorts of other growing pains as well, such as having to deal with
external users, running mailing lists, portability, managing complexity and
other issues. Some of these are discussed in \cite{BH13}. The point simply is
that there is no need to worry too much about testing tools for programs of
this size: care about writing tests from the beginning and if the number of
tests becomes too unwieldy, there will be professional-grade tools you can
switch to that can manage this for you. In the meantime, a little shell script
such as the following will do just fine:
\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]{tests/compare.sh}
If you call this script, say, \texttt{tests/compare.sh}, then running it using
the command \texttt{cd tests/ ; bash -e compare.sh} should produce output like
this:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
 .    testcase-1.json
 .    testcase-2.json
 .    testcase-3.json
 .    testcase-4.json
 .    testcase-5.json
\end{lstlisting}
On the other hand, if your output looks like this
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
 .    testcase-1.json
 .    testcase-2.json
 .    testcase-3.json
 X    testcase-4.json
 .    testcase-5.json
\end{lstlisting}
then you know that you just broke some functionality that
\texttt{testcase-4.json} checks. Given how easy it is to run this script, you
want to run this script often to find new bugs as early as possible.

@@

From experience, it turns out that this is not often a real problem. The vast
majority of patches concern just small, isolated pieces of the software and in
those cases, only one or a few testcases will start failing. It is then easy
to identify the simplest one of these (assuming that they were given useful
names, see above in Section~\ref{sec:many-tests}), and that is the one we
will then use for debugging. If one breaks fundamental functionality somewhere
in the core of the software, a large number of tests will start to fail
but in those cases you will either be able to quickly find out what is
happening, or be able to debug the problem by choosing \textit{any} simple
test that was among those that broke. In summary, from a practical
perspective, we have found that it is not usually a problem to have testcases
that check overlapping parts of the software. Rather, we have come to the
conclusion that having as many tests as possible, regardless of overlap.

@@

At the base of the problem is the fact that floating point numbers do not
enjoy the same mathematical properties as integers (or, in fact, as real
numbers). For example, if you take into account round-off, then $(a+b)+c$ is in
general \textit{not} the same as $a+(b+c)$ or $(a+c)+b$. You can try this out
with $a=1.0, b=-1.0, c=1e-20$: in double precision, you will get
$(a+b)+c=1e-20, a+(b+c)=(a+c)+b=0$. This matters because if you make changes
to the code base, upgrade the compiler, switch from one processor or machine
to another, or change compiler optimization level, floating point operations
will be executed in different order and the results will be different at the
level of roundoff. In other words, your tests will fail even though
nothing fundamentally changed.

@@

A second area that creates problems is that codes often try to compute zeros, but
because of roundoff what is actually produced are numbers like
4.863e-19. A human recognizes that this is a zero and, consequently, equal to
2.489e-19 if all of the numbers around it are on the order of one. But the
\texttt{diff} command will complain, and consequently signal a difference.

The solution to these problems that we have ultimately come to use
is a program called \texttt{numdiff}.%
\footnote{See \url{http://www.nongnu.org/numdiff/}.}
\texttt{numdiff} works essentially like
\texttt{diff} but it recognizes if the text it compares in two files is
actually a number. If two numbers differ by less than some relative or
absolute tolerance, then they are considered equal even if the characters by
which they are represented in an output file are different. For example, you
would replace the \texttt{diff} command used previously by the
following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
10^{-6}$ or $|a-b|\le 10^{-8} \min\{|a|,|b|\}$:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
\end{lstlisting}
Using this trick, numbers that are written into the output file
\texttt{testcase-1.output.new} on one machine with slight differences to the
ones originally saved in \texttt{testcase-1.output} will still be considered
the same, and the comparison will succeed.

@@

When we talk to students who just starting writing their own codes, then
they typically understand the importance of testing in some abstract
sense. But almost all of them have two fundamental misconceptions: First, that
their codes are either written well enough or will not be changed in ways
fundamental enough to suggest that these changes might break existing
functionality. And second, that doing adequate testing is just too hard and
cumbersome to be a justifiable expense of time.

The first of these two points is empirically false. Only bad programmers
expect that they will not break existing functionality, no matter how trivial
a change might be. All experienced programmers will know this to be wrong. The
reason is that over the lifetime of a scientific code -- say, over the course of
the 3 years a graduate student does research, and maybe even beyond this if
the code is handed on to the next student, or if you go on to be a postdoc --
a code starts out small but is constantly developed into something much larger
with much more functionality in all sorts of directions. You're not going to
create one essential version of the code that you can verify to be correct and
then just round off some edges; you're in fact going to replace the linear by
a nonlinear model, replace the direct solver by an iterative one, add terms to
the equation, and make other very fundamental changes that will require you to
edit all throughout your code. There is no other way than systematic testing
to ensure that the computation you ran today will still run tomorrow. And you
will have to run it again tomorrow: your adviser or boss may ask you to redo a
computation with some slight changes, a reviewer of a paper might ask to do
additional simulations, you may have forgotten the exact parameters for a data
set you want to use in your thesis and will have to redo it to verify,
etc. You may think that once you have created the data 
for some particular simulation, that the issue is done and will not have to
repeated -- i.e., that you might in fact just throw away the code -- but this
is not typically true.
